\title{Algorithm to find number of derangements of length N with K good swaps}
\author{
    Christopher He
}
% \date{\today}

\documentclass[12pt]{article}
\usepackage{amsmath}

\begin{document}
\maketitle

% \begin{abstract}
% This is the paper's abstract \ldots
% \end{abstract}

\section{Introduction}

We consider a derangement of size $N$ to be an array $D$ that is a permutation of ${0, …, N - 1}$
where no element is in its correct sorted position.
That is, for all $i, 0 <= i < N, D[i] \neq i$. Note that in our definition of derangement,
all elements are distinct.

% \paragraph{Outline}
% The remainder of this article is organized as follows.
% Section~\ref{previous work} gives account of previous work.
% Our new and exciting results are described in Section~\ref{results}.
% Finally, Section~\ref{conclusions} gives the conclusions.

A good swap in a derangement $D$ of size $N$ is a pair $(i, j), 0 <= i < j < N$
such that $D[i] = j$ or $D[j] = i$.
% In English, a swap is good if it places at least one element in its correct sorted position.

Let s be the number of good swaps in a derangement D of size N.
First observe that for every element $x$ in $D$, there is at most one good swap that places $x$ in its
correct position, so $s <= N$. Also, $s > \lfloor N / 2 \rfloor$, this follows from the fact that a good swap,
 in the best case, places two elements in their correct positions (i.e $D[i] = j$ AND $D[j] = i$).

So we know that $ \lfloor N/2 \rfloor < s <= N$. using these bounds, we can show that the expected number of random
swaps to sort D is between $N^2/8$ and $N^2/2$. To get a more exact running time, it could be
helpful to find the expected number of good swaps. This is defined as

\[ E[s] = \sum_{k = \lfloor N/2 \rfloor}^{N} p_k \times k \]

Where $p_k$ is the probability of a derangement having $k$ good swaps. We can define this as

\[ \frac{\text{number of derangements of size $N$ with $k$ good swaps}}{\text{total number of derangements of size $N$}} \]

It can be shown that the number of derangements of size N is exactly $\lfloor \frac{N! + 1}{e} \rfloor$.
But how many derangements of size $N$ have $K$ good swaps?


\section{Model as Graph}
Given a derangement $D$ of size $N$ we construct a directed graph $G(V, E)$ where $V = \{0, ..., N - 1\}$
and $(i, j) \in E$ if $D[i] = j$. First observe that there are no self-loops in this graph because in 
a derangement there is no element in its correct position. Secondly, observe that each vertex $i$ has an outdegree
of exactly $1$, that is, $D[i] = j$ for some $j$ and if $D[i] = j$ and $D[i] = k$ then $j = k$. A similar argument shows that each
vertex has an indegree of $1$. With these observations, the graph must be a connected cycle graph, or if
disconnected, decomposed into multiple cycle graphs each with 2 or more vertices.
Let's call a graph with these properties a \textbf{derangement graph}.

The number of good swaps in $D$ is almost the number of edges in its derangement graph, except if we have a cycle of $2$ vertices,
that represents $1$ good swap, whereas if we have a cycle of $m$ vertices where $m > 2$ then that cycle contributes
$m$ good swaps ($m$ vertices implies $m$ edges, each being an instance where $D[i] = j$ so one of $(i, j)$ or 
$(j, i)$ is a good swap).

We then rephrase our original problem to the following: how many derangement graphs with
vertices $\{0, ..., N - 1\}$ are there with $k$ good swaps?

\section{Recurrence}
We start with a naive recurrence:
Let T(N, k) be the number of derangement graphs with vertices $\{0, ..., N - 1\}$ with $k$ good swaps where
$k <= N$.
We define a recurrence for T(N, k) based on the following idea:

Pick $i$ vertices from the $N$ vertices, where $i > 1$. With these $i$ vertices, we will make a cycle of length $i$.

If we make a cycle of length $2$, that contributes $1$ good swap.
We would like the rest of the $N - 2$ vertices to contribute $k - 1$
good swaps. Assume we have $T(N - 2, k - 1)$

If we make a cycle of length $i$ where $i > 2$, that contributes
$i$ good swaps. Assume we have $T(N - i, k - i)$.

Let $C_i$ be the number of cycles we can make of length $i$.
There are ${N \choose i}$ ways to pick $i$ vertices. With these $i$ vertices,
there are $(i-1)!$ cycles we can make (fix a starting vertex, there
are $(i - 1)$ choices for the second vertex in the cycle,
then $(i - 2)$, and so on). 

So $C_i = {N \choose i} \times (i - 1)!$.

Putting this all together, we have the following recurrence:
\[ T(N, k) =
C_2 \times T(N - 2, k - 1) +
\sum_{i = 2}^{k}{C_i \times T(N-i, k-i)} \]

This recurrence is naive because of the following observation:
Take for example a graph of 11 vertices. Let's say we split this
graph into cycles of lengths 4, 3, 2, and 2. We counted this
arrangement $4!$ times since there are that many ways to permute these
cycles. Thus we define another recurrence $T^\prime(N, k, s)$, the number
of ways to make a derangement graph of N vertices with k good swaps
with s connected components. The naive reccurence is slightly modified:

\[ T^\prime(N, k, s) =
C_2 \times T^\prime(N - 2, k - 1, s-1) +
\sum_{i = 2}^{k}{C_i \times T^\prime(N-i, k-i, s-1)} \]
Thus our final solution is
\[ T(N, K) = \sum_{s = 1}^{N}{\frac{T^\prime(N, K, s)}{s!}} \]

\section{Conclusions}\label{conclusions}
Running this algorithm shows that the expected number of good swaps
in a derangement of size $N$ is almost exactly $N - 0.5$.
One way we can use this fact is to derive a more exact expected running time of the
randomly swapping algorithm. Another way is, since we have the expected number of
good swaps, we have the expected number of swaps that aren't good.

% \bibliographystyle{abbrv}
% \bibliography{main}

\end{document}
This is never printed

